---
title: "Shooting project"
author: "H.Lm"
date: "2024-02-12"
output:
  html_document: default
  pdf_document: default
---
## I want to predict whether or not a victim is likely to die in a shooting in NYC
### Explanation of my approach
I see that in the data we have location-based data, time data, and information about victim and perpetrator.
I'll run a model with the human-readable factors to see what stands out as significantly contribution to likelihood of shooting resulting in murder.  There are three steps to that

* Clean the data, looking at exploratory trends
* Build the model
* Data analysis


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(stats)
```

## Read in data
#### R packages used: tidyverse, lubridate, stats
``` {r echo = TRUE}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
data <- read.csv(url)
dim(data)
```

## Step 1: Data build the DATA for a model that predicts whether or not someone will die in a shooting
### Step 1a: add in derived variables for date
``` {r echo = TRUE}
# Each row is a single shooting
# The columns are only the values that are human-interpretable

# create derived year, month, hour
model_data <- data %>% 
  mutate(
    OCCUR_DATE = mdy(OCCUR_DATE),
    OCCUR_TIME = hms(OCCUR_TIME),
    year = lubridate::year(OCCUR_DATE),
    month = lubridate::month(OCCUR_DATE),
    hour = hour(OCCUR_TIME)
  ) %>%
   mutate(
    hour = case_when(
      hour >= 5  & hour <= 11  ~ "Morning",
      hour >= 12 & hour <= 16  ~ "Afternoon",
      hour >= 17 & hour <= 20  ~ "Evening",
      hour >= 21 | hour <= 4   ~ "Night"
    )
  ) %>%
  mutate(
    season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% c(3, 4, 5) ~ "Spring",
      month %in% c(6, 7, 8) ~ "Summer",
      month %in% c(9, 10, 11) ~ "Autumn"
    )
  ) %>%
  mutate(precinct_cat = as.factor(PRECINCT))
```

### Step 1b: remove unneeded columns
``` {r echo = TRUE}

# create derived year, month, hour
model_data <- model_data %>% 
  select(OCCUR_DATE, year, season, hour,
         # month, hour, 
         BORO, precinct_cat,
         # LOCATION_DESC, LOC_OF_OCCUR_DESC, LOC_CLASSFCTN_DESC,
         PERP_AGE_GROUP, PERP_SEX, PERP_RACE,
         VIC_AGE_GROUP, VIC_SEX, VIC_RACE,
         STATISTICAL_MURDER_FLAG
  )
```

### Step 1c: placeholder for cleanup found from later steps
``` {r echo = TRUE}

# first transform STATISTICAL_MURDER_FLAG to be 0 or 1
model_data$STATISTICAL_MURDER_FLAG <- as.numeric(as.logical(model_data$STATISTICAL_MURDER_FLAG))

# make a categorical flag for murder
model_data <- model_data %>%
  mutate(murder_flag = ifelse(STATISTICAL_MURDER_FLAG ==0, "NON-FATAL","FATAL"))

# let's remove that weird single data point of entry error
model_data <- model_data %>%
  filter(VIC_AGE_GROUP != '1022') %>%
  filter(!PERP_AGE_GROUP %in% c('1022','940','224','1020'))

# let's combine all the unknown & blanks into one category
model_data <- model_data %>%
  mutate(PERP_RACE = replace(PERP_RACE, PERP_RACE %in% c("", "(null)", "UNKNOWN"), "Unknown")) %>%
  mutate(PERP_AGE_GROUP = replace(PERP_AGE_GROUP, PERP_AGE_GROUP %in% c("", "(null)", "UNKNOWN"), "Unknown")) %>%
  mutate(PERP_SEX = replace(PERP_SEX, PERP_SEX %in% c("", "(null)", "UNKNOWN", "U"), "Unknown"))


```

## Step 1d: Always check time trends
``` {r echo = FALSE}
monthly_counts <- model_data %>%
  group_by(year = year(OCCUR_DATE), month = month(OCCUR_DATE), BORO) %>%
  summarise(count = n())

# Create the plot
ggplot(monthly_counts, aes(x = as.Date(paste(year, month, "01", sep = "-")), y = count, group = BORO, color = BORO)) +
  geom_line() +
  labs(title = "Trend of Shootings Count Over Time by Borough (Aggregated by Month)",
       x = "Date", y = "Count") +
  theme_minimal()

# Aggregate the data by year and borough
yearly_counts <- model_data %>%
  group_by(year = year(OCCUR_DATE), BORO) %>%
  summarise(count = n())

# Create the plot
ggplot(yearly_counts, aes(x = year, y = count, group = BORO, color = BORO)) +
  geom_line() +
  geom_point(size = 3) +
  geom_text(aes(label = year), hjust = 1.5, vjust = 0) +
  labs(title = "Trend of Shootings Count Over Time by Borough (Aggregated by Year)",
       x = "Year", y = "Count") +
  theme_minimal()

# Aggregate the data by year and fatality and borough
yearly_counts <- model_data %>%
  group_by(year = year(OCCUR_DATE), BORO, murder_flag) %>%
  summarise(count = n())

# Create the stacked bar chart
ggplot(yearly_counts, aes(x = year, y = count, fill = murder_flag)) +
  geom_bar(stat = "identity") +
  labs(title = "Stacked Bar Chart of Shootings Fatalities Over Time by Borough",
       x = "Year", y = "Count") +
  theme_minimal()
```
#### OBSERVATIONS

* Clearly there are both seasonal trends and long-term trends over time
* I guess the pandemic made people angry.
* However, the rates of fatal shootings do not seem to vary by year, so we will not consider this time trend in the analysis



## Step 2: Run the model
``` {r echo = TRUE}

logistic_model <- glm(STATISTICAL_MURDER_FLAG ~ .-OCCUR_DATE-year-murder_flag-precinct_cat, data = model_data, family = binomial)

# Print a summary of the model
summary(logistic_model)
```

#### We can see that the following factors seem to contribute to likelihood of a murder from shooting:
* Season (summer shooting is less likely to result in murder)
* Time of day (morning shooting is more likely to result in murder)
* Location (Manhattan seems to have less fatal shootings)
* Perpetrator characteristics
** Age group: shooters older than 18 tend to result in fatalities, while unknown age result in fewer fatalities
** Sex: males tended to result in fewer fatalities, while sex unknown resulted in more <- most unintuitive
* Victim characteristics: only victim age > 18 tended to result in fatalities

Let's look into each of these!

## Step 3: Look into strangely significant variables from the model

### Step 3a: Victim age groups stand out: all defined age groups were significant except <18 & unknown
NOTE: I'm showing the code for this block, but all charts below are made similarly and turning echo = FALSE for them.
``` {r echo = TRUE}
# histogram victim age group
model_data %>%
  ggplot(aes(x=VIC_AGE_GROUP, fill = murder_flag)) +
  geom_bar(position = "stack") +
  labs(x = "victim age group", y = "count", fill = "murder or no")

# look at the actual % fatality
summary_vic_age <- model_data %>%
  group_by(VIC_AGE_GROUP, murder_flag) %>%
  summarise(count = n(),.groups="drop") %>%
  group_by(VIC_AGE_GROUP) %>%
  mutate(percent_fatal = count/sum(count)*100) %>%
  filter(murder_flag != "NON-FATAL") %>%
  mutate(total_shootings = round(count*100/percent_fatal)) %>%
  select(-murder_flag,-count)

summary_vic_age %>%
  ggplot(aes(x = VIC_AGE_GROUP)) +
     geom_bar(aes(y = total_shootings), stat = "identity", fill = "skyblue", alpha = 0.5) +
     labs(y = "Total Shootings") +
    geom_line(aes(y = percent_fatal*max(total_shootings)/max(percent_fatal), group = 1), 
                   color = "red", linewidth = 1) +
      scale_y_continuous(
        sec.axis = sec_axis(
          ~ . * max(summary_vic_age$percent_fatal) / max(summary_vic_age$total_shootings),
          name = "Percent Fatal"
        )
      ) +
      labs(x = "Victim Age")
```

#### OBSERVATIONS

* Likelihood of dying from a shooting increases linearly with age.
* Vast majority of shootings happen for 18-44 year olds.

### Step 3b: Perpetrator race does NOT stand out, but let's make the same kind of histogram as above anyways since I coded it up by chance
``` {r echo = FALSE}

# histogram perp race
model_data %>%
  ggplot(aes(x=PERP_RACE, fill = murder_flag)) +
  geom_bar(position = "stack") +
  labs(x = "perp race", y = "count", fill = "murder or no") +
  coord_flip()

# look at the actual % fatality
summary_perp_race <- model_data %>%
  group_by(PERP_RACE, murder_flag) %>%
  summarise(count = n(),.groups="drop") %>%
  group_by(PERP_RACE) %>%
  mutate(percent_fatal = count/sum(count)*100) %>%
  filter(murder_flag != "NON-FATAL") %>%
  mutate(total_shootings = round(count*100/percent_fatal)) %>%
  select(-murder_flag,-count)

summary_perp_race %>%
  ggplot(aes(x = PERP_RACE)) +
     geom_bar(aes(y = total_shootings), stat = "identity", fill = "skyblue", alpha = 0.5) +
     labs(y = "Total Shootings") +
    geom_line(aes(y = percent_fatal*max(total_shootings)/max(percent_fatal), group = 1), 
                   color = "red", linewidth = 1) +
      scale_y_continuous(
        sec.axis = sec_axis(
          ~ . * max(summary_perp_race$percent_fatal) / max(summary_perp_race$total_shootings),
          name = "Percent Fatal"
        )
      ) +
      labs(x = "Perp Race") +
  coord_flip()
```

#### OBSERVATIONS

* There are trends in total counts of perpetrator races
* It seems like fatality rate inversely proportional with relative count of shooting by race
* While these trends are interesting, race is NOT statistically significant as a factor in the model, indicating there are likely correlations between this variable and other variables


### Step 3c: Perpetrator age stands out
``` {r echo = FALSE}

# histogram perp age
model_data %>%
  ggplot(aes(x=PERP_AGE_GROUP, fill = murder_flag)) +
  geom_bar(position = "stack") +
  labs(x = "perpetrator age group", y = "count", fill = "murder or no")

  
# look at the actual % fatality
summary_perp_age <- model_data %>%
  group_by(PERP_AGE_GROUP, murder_flag) %>%
  summarise(count = n(),.groups="drop") %>%
  group_by(PERP_AGE_GROUP) %>%
  mutate(percent_fatal = count/sum(count)*100) %>%
  filter(murder_flag != "NON-FATAL") %>%
  mutate(total_shootings = round(count*100/percent_fatal)) %>%
  select(-murder_flag,-count)

summary_perp_age %>%
  ggplot(aes(x = PERP_AGE_GROUP)) +
     geom_bar(aes(y = total_shootings), stat = "identity", fill = "skyblue", alpha = 0.5) +
     labs(y = "Total Shootings") +
    geom_line(aes(y = percent_fatal*max(total_shootings)/max(percent_fatal), group = 1), 
                   color = "red", linewidth = 1) +
      scale_y_continuous(
        sec.axis = sec_axis(
          ~ . * max(summary_perp_age$percent_fatal) / max(summary_perp_age$total_shootings),
          name = "Percent Fatal"
        )
      ) +
      labs(x = "Perp Age")

```

#### OBSERVATIONS

* The histogram looks similar to victim age with two exceptions
  * HUGE stack of unknown age of perpetrators
  * Order of 18-24 than 25-44 proportion of shootings flipped
* The fatality order mirrors the victim age exactly: don't get shot by a senior

### Step 3d: Perpetrator sex stands out
``` {r echo = FALSE}

# histogram perp sex
model_data %>%
  ggplot(aes(x=PERP_SEX, fill = murder_flag)) +
  geom_bar(position = "stack") +
  labs(x = "perpetrator sex", y = "count", fill = "murder or no")

# look at the actual % fatality
summary_perp_sex <- model_data %>%
  group_by(PERP_SEX, murder_flag) %>%
  summarise(count = n(),.groups="drop") %>%
  group_by(PERP_SEX) %>%
  mutate(percent_fatal = count/sum(count)*100) %>%
  filter(murder_flag != "NON-FATAL") %>%
  mutate(total_shootings = round(count*100/percent_fatal)) %>%
  select(-murder_flag,-count)

summary_perp_sex %>%
  ggplot(aes(x = PERP_SEX)) +
     geom_bar(aes(y = total_shootings), stat = "identity", fill = "skyblue", alpha = 0.5) +
     labs(y = "Total Shootings") +
    geom_line(aes(y = percent_fatal*max(summary_perp_sex$total_shootings)/max(summary_perp_sex$percent_fatal), group = 1), 
                   color = "red", linewidth = 1) +
      scale_y_continuous(
        sec.axis = sec_axis(
          ~ . * max(summary_perp_sex$percent_fatal) / max(summary_perp_sex$total_shootings),
          name = "Percent Fatal"
        )
      ) +
      labs(x = "Perp Sex")
```

#### OBSERVATIONS: SIMPSON'S PARADOX!

* Obviously way more male and unknown sex perpetrators (ban men from owning guns is my personal bias)
* The order of raw % fatality of perpetrators by sex in descending order: F > M > unknown
  * This is also fairly intuitive as the groups less likely to be perpetrators tend to be more likely to shoot to kill
  * However, this is OPPOSITE what the model's effects indicate: that an unknown sex tends to result in HIGHER likelihood of a fatal shooting.
  * For this analysis to be complete, this example of Simpson's paradox (Simpson's paradox occurs when a trend or relationship that appears in different groups disappears or reverses when the groups are combined) must be explained by some other variable that correlates and subdivides the data in more ways

### Step 3e: Borough analysis: Manhattan stands out
``` {r echo = FALSE}

# histogram borough
model_data %>%
  ggplot(aes(x=BORO, fill = murder_flag)) +
  geom_bar(position = "stack") +
  labs(x = "borough", y = "count", fill = "murder or no")

# show actual % fatalities
summary_boro <- model_data %>%
  group_by(BORO, murder_flag) %>%
  summarise(count = n(),.groups="drop")

summary_boro <- summary_boro %>%
  group_by(BORO) %>%
  mutate(percent_fatal = count/sum(count)*100) %>%
  filter(murder_flag != "NON-FATAL") %>%
  mutate(total_shootings = round(count*100/percent_fatal)) %>%
  select(-murder_flag,-count)

summary_boro %>%
  ggplot(aes(x = BORO)) +
     geom_bar(aes(y = total_shootings), stat = "identity", fill = "skyblue", alpha = 0.5) +
     labs(y = "Total Shootings") +
    geom_line(aes(y = percent_fatal*max(summary_boro$total_shootings)/max(summary_boro$percent_fatal), group = 1), 
                   color = "red", linewidth = 1) +
      scale_y_continuous(
        sec.axis = sec_axis(
          ~ . * max(summary_boro$percent_fatal) / max(summary_boro$total_shootings),
          name = "Percent Fatal"
        )
      ) +
      labs(x = "Borough")

```

#### OBSERVATIONS

* Possibly, Manhattan is better-policed.  I have no other explanations.

### Step 3f: Time of day analysis
``` {r echo = FALSE}
# histogram time of day group
model_data %>%
  ggplot(aes(x=hour, fill = murder_flag)) +
  geom_bar(position = "stack") +
  labs(x = "time of day", y = "count", fill = "murder or no")

# look at the actual % fatality
summary_hour <- model_data %>%
  group_by(hour, murder_flag) %>%
  summarise(count = n(),.groups="drop") %>%
  group_by(hour) %>%
  mutate(percent_fatal = count/sum(count)*100) %>%
  filter(murder_flag != "NON-FATAL") %>%
  mutate(total_shootings = round(count*100/percent_fatal)) %>%
  select(-murder_flag,-count)

summary_hour %>%
  ggplot(aes(x = hour)) +
     geom_bar(aes(y = total_shootings), stat = "identity", fill = "skyblue", alpha = 0.5) +
     labs(y = "Total Shootings") +
    geom_line(aes(y = percent_fatal*max(total_shootings)/max(percent_fatal), group = 1), 
                   color = "red", linewidth = 1) +
      scale_y_continuous(
        sec.axis = sec_axis(
          ~ . * max(summary_hour$percent_fatal) / max(summary_hour$total_shootings),
          name = "Percent Fatal"
        )
      ) +
      labs(x = "Time of day")
```

#### OBSERVATIONS

* Most shootings at night, lowest accuracy at night (it's dark?)
* Morning spike because maybe shooters are not morning people


### Step 3g: Seasonality analysis
``` {r echo = FALSE}
# histogram season group
model_data %>%
  ggplot(aes(x=season, fill = murder_flag)) +
  geom_bar(position = "stack") +
  labs(x = "season", y = "count", fill = "murder or no")

# look at the actual % fatality
summary_season <- model_data %>%
  group_by(season, murder_flag) %>%
  summarise(count = n(),.groups="drop") %>%
  group_by(season) %>%
  mutate(percent_fatal = count/sum(count)*100) %>%
  filter(murder_flag != "NON-FATAL") %>%
  mutate(total_shootings = round(count*100/percent_fatal)) %>%
  select(-murder_flag,-count)

summary_season %>%
  ggplot(aes(x = season)) +
     geom_bar(aes(y = total_shootings), stat = "identity", fill = "skyblue", alpha = 0.5) +
     labs(y = "Total Shootings") +
    geom_line(aes(y = percent_fatal*max(total_shootings)/max(percent_fatal), group = 1), 
                   color = "red", linewidth = 1) +
      scale_y_continuous(
        sec.axis = sec_axis(
          ~ . * max(summary_season$percent_fatal) / max(summary_season$total_shootings),
          name = "Percent Fatal"
        )
      ) +
      labs(x = "Season")
```

#### OBSERVATIONS

* Most shootings in summer, lowest accuracy in summer
* Winter has way fewer shootings, maybe fatalities are higher indoors?


# Conclusions

### From our analysis, we were able to validate some of the model important variables via intuition

* Typically, young victims are more likely to recover because they are healthier and physically recover more quickly than adults.
* Unknown age could result in fewer fataliites because less is known about the shooter.  It could be a stray bullet which wasn't aimed.

### Things worth exploring more

* Manhattan is one of the safer boroughs, so if by chance a shooting happens, it's not likely to be fatal, or maybe the policing is better and therefore fewer fatal shots happen
* Time of day was interesting, as was seasonality.  There are definitely trends in number of shootings, but why morning shootings are more often fatal I cannot think of a good reason for. Night shootings had lower accuracy because if it's happening outside it's dark and harder to aim.


## Sources of Bias
- I am assuming a lot about what “unknown” implies - that they perpetrator was not caught and therefore was further away from the victim.  Perpetrators who are unknown could more likely be biased to a category and I have no data about this.  My bias is that having unknown characters for a perpetrator is related to bad policing (because the shooter is not caught).
- If you try including precinct in the model, the precincts in Brooklyn tended to be associated with higher fatality rates than the precincts in other areas.  My bias if I were to explore this is that there is definitely some kind of connection with demographics and quality of policing.  This is an area I would want to explore in the future.
- This is confirmed by the fact that in 2020, the increase in shootings spiked in Brooklyn more than other areas.
- To remove this source of bias, I would want to look at the multicollinearity between the variables that I included in my model.


